---
title: Ft Riley butterfly surveys
description: |
  Preliminary assessment of IMAP data from 2021 - Regal
author:
  - name: Rexstad 
    url: 
    affiliation: CREEM Univ St Andrews
    affiliation_url: https://www.creem.st-andrews.ac.uk/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 1
    toc_float: true
---

This exercises takes a first look at the 2021 Regal butterfly survey.  We take an initial look at the distribution of distances and fit models to the data. The models examine whether detection functions differ between the four sampling bouts of the season.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
solution <- TRUE
```

Load necessary packages

```{r pkgs}
library(readxl)
library(Distance)
library(vioplot)
library(knitr)
library(kableExtra)
```

# Acquire data

```{r data}
spreadsheet.regal <- "IMAP Regal Data_ALL.xlsx"
regal <- as.data.frame(read_xlsx(path=spreadsheet.regal, sheet="Sheet1"))
```

# Some renaming of fields

The `Distance` package is quite specific about the names associated with some elements of the data. For example, the field `observer` is specific to surveys that are conducted using multiple simultaneous observers to estimate detectability on the transect.  As this is not the case for this survey, rename that field to `obsname` to avoid difficulties.

```{r rename}
regal$Region.Label <- regal$Region
regal$Area <- regal$`Area (ha)`
regal$Sample.Label <- regal$`Transect ID`
regal$Effort <- regal$`Transect Length (m)`
regal$distance <- regal$`Distance (m)`
names(regal)[9] <- "obsname"
regal$sample.bout <- regal$`Sample Bout`
```

Specify units used for various measurements: perpendicular distances, transect length and density estimates.

```{r units}
bflyunits <- convert_units("meter", "meter", "hectare")
```

# Exploratory

Use a large number of bins to gain fine-scale information about distribution of detection distances.

```{r explore, eval=solution}
hist(regal$distance, nc=50, xlab="Perpendicular distance", 
     main="Regal detections 2021 pooled across visits")
```

Distribution of detection distances by observer and by sampling bout

```{r moreEDA, eval=solution}
obs <- unique(regal$obsname)
vioplot(distance~obsname, data=regal, main="Detection distance distribution by observer")
vioplot(distance~sample.bout, data=regal, main="Detection distance distribution by bout")
```
```{r eda, eval=solution}
b <- vector(mode="integer", length=4)
b[1] <- sum(!is.na(regal$distance[regal$sample.bout==1]))
b[2] <- sum(!is.na(regal$distance[regal$sample.bout==2]))
b[3] <- sum(!is.na(regal$distance[regal$sample.bout==3]))
b[4] <- sum(!is.na(regal$distance[regal$sample.bout==4]))
kable(data.frame(numdetect=b), row.names = TRUE, caption="Detections by sampling bout.") %>%
  kable_paper(full_width=FALSE)
```

# Begin analysis

Select a truncation distance based upon visual inspection of perpendicular distance histogram.  After fitting, perform goodness of fit tests and capture GOF test output without displaying the QQ plot.

```{r firstfit, eval=solution}
w <- vector("numeric", 2)
p <- vector("numeric", 2)
mytrunc <- 45
monhn <- ds(data=regal, key="hn", convert.units = bflyunits,
            formula=~as.factor(sample.bout), truncation = mytrunc)
hnfit <- gof_ds(monhn, plot=FALSE)$dsgof$CvM
w[1] <- round(hnfit$W, 4)
p[1] <- round(hnfit$p, 5)
monhr <- ds(data=regal, key="hr", convert.units = bflyunits,
            formula=~as.factor(sample.bout), truncation = mytrunc)
hrfit <- gof_ds(monhr, plot=FALSE)$dsgof$CvM
w[2] <- round(hrfit$W, 4)
p[2] <- round(hrfit$p, 5)
kable(data.frame(w, p, row.names = c("Halfnorm", "Hazard")), row.names=TRUE,
      caption="Cramer-von Mises goodness of fit for two key function models.")
```

Based upon preliminary modelling, focus upon the hazard rate detection function.  Examine the details of detection function modelling.  Furthermore, as we are combining the sampling bouts, but want an overall estimate of density across all bouts, use `dht2`.  This function properly estimates uncertainty across the replicate surveys using the `stratification=` argument.

```{r estimates, eval=solution}
summary(monhr$ddf)
monhr.ests <- dht2(ddf=monhr, flatfile=regal,
                  strat_formula = ~as.factor(sample.bout), 
                  convert_units = bflyunits,
                  stratification = "replicate")
print(monhr.ests, report="density")
```

Examine session-specific detection function differences.  Compare visual output with parameter estimates shown in previous output.

```{r hrboutplot, fig.cap="Hazard rate with bout-specific detection function.", eval=solution}
visitplot <- function(dsobject) {
# plots visit-specific detection functions
  plot(dsobject, showpoints=FALSE, nc=40,
       main="Detection function with visit as covariate")
  add_df_covar_line(dsobject, data=data.frame(sample.bout=as.factor(1)), 
                    lty=1, col="blue")
  add_df_covar_line(dsobject, data=data.frame(sample.bout=as.factor(2)), 
                    lty=1, col="green")
  add_df_covar_line(dsobject, data=data.frame(sample.bout=as.factor(3)),
                    lty=1, col="red")
  add_df_covar_line(dsobject, data=data.frame(sample.bout=as.factor(4)),
                    lty=1, col="purple")
  legend("topright", lty=1, legend=c("Visit 1", "Visit 2", "Visit 3", "Visit 4"),
         col=c("blue", "green", "red"))
}
visitplot(monhr)
```

# Pooled density analysis

For completeness, fit detection function models using both key functions combining data across all bouts.  Note that `Effort` is recalculated when data are combined across sessions.

```{r pooled, eval=solution}
regal$Effort <- regal$Effort * length(unique(regal$sample.bout))
m.pool.hn <- ds(regal, key="hn", convert.units = bflyunits, truncation = mytrunc)
m.pool.hr <- ds(regal, key="hr", convert.units = bflyunits, truncation = mytrunc)
kable(summarize_ds_models(m.pool.hn, m.pool.hr)[,2:7], digits=4, row.names=FALSE)
plot(m.pool.hr, nc=58, main="2021 Regal data with a single\nfitted detection function.")
summary(m.pool.hr)
```

#  Does detectability differ between sampling bouts?

AIC should provide an answer to this

```{r aic, eval=solution}
kable(summarize_ds_models(m.pool.hn, m.pool.hr, monhn, monhr)[,2:7], digits=4, row.names=FALSE)
```

# Comments

What conclusions can we draw from basic analysis of the regal data set?

## Spike on the transect
Note the large proportion of detections in the first meter `r round(100*sum(regal$distance<=1, na.rm=TRUE)/sum(regal$distance<=mytrunc, na.rm=TRUE),3)`\% of all detections are in the first meter (the first `r round(100*1/45, 1)`% of the sampled strip).  This leads to a spike close to zero, which is difficult to fit.  There are three consequences:

- contrary to the usual situation with line transect data where 25% of the uncertainty in the density estimate is attributable to uncertainty in the detection function, here `r round(100*(as.numeric(summary(m.pool.hr)$ds$average.p.se / summary(m.pool.hr)$ds$average.p))^2/m.pool.hr$dht$individuals$D$cv^2,1)`\% of the uncertainty comes from the detection function.

## Comparison of overall estimates

Bout-specific estimates of density are highly variably, with few detections in both June and September. Coefficients of variation for the July and August estimates are both >0.3.  
- The point estimate of $\hat{D}$ for the analysis of the pooled data is `r round(m.pool.hr$dht$individuals$D$Estimate,3)` butterflies per ha.

